Rotation and Lithium Confirmation of a 500 Parsec Halo for the Open Cluster NGC\,2516\footnote{This is paper 3 of the Cluster Difference Imaging Photometric Survey.}

### 2021-04-26 cluster_paper/maintable_header.tex
update ms

%% This is the title of the table.
\tablecaption{Rotation periods and lithium equivalent widths for
candidate NGC\,2516 members.}
\label{tab:maintable}

@@

             \texttt{source\_id} & 5290824530952816896 &                                         Gaia DR2 source identifier. \\
       \texttt{n\_cdips\_sector} &                   3 &                     Number of TESS sectors with CDIPS light curves. \\
                 \texttt{period} &            1.297662 &                                           Lomb-Scargle best period. \\
                 \texttt{lspval} &            0.171629 &                     Lomb-Scargle periodogram value for best period. \\
                 \texttt{nequal} &                 0.0 &          Number of stars brighter than the target in TESS aperture. \\
                 \texttt{nclose} &                 1.0 &            Number of stars with $\Delta T > 1.25$ in TESS aperture. \\
                 \texttt{nfaint} &                 1.0 &             Number of stars with $\Delta T > 2.5$ in TESS aperture. \\
             \texttt{spdmperiod} &            1.297662 &                                       Stellingwerf PDM best period. \\
                \texttt{spdmval} &            0.826804 &                 Stellingwerf PDM periodogram value for best period. \\
             \texttt{ref\_epoch} &              2015.5 &                                         Reference epoch for RA/dec. \\
                     \texttt{ra} &          120.125821 &                                           Gaia DR2 right ascension. \\
                    \texttt{dec} &          -60.522668 &                                               Gaia DR2 declination. \\
               \texttt{parallax} &            1.980389 &                                                  Gaia DR2 parallax. \\
        \texttt{parallax\_error} &            0.031363 &                                      Gaia DR2 parallax uncertainty. \\
                   \texttt{pmra} &           -4.480769 &                    Gaia DR2 proper motion $\mu_\alpha \cos \delta$. \\
                  \texttt{pmdec} &            8.203306 &                                Gaia DR2 proper motion $\mu_\delta$. \\
     \texttt{phot\_g\_mean\_mag} &           12.958735 &                                             Gaia DR2 $G$ magnitude. \\
    \texttt{phot\_bp\_mean\_mag} &           13.309761 &                                 Gaia DR2 $G_\mathrm{BP}$ magnitude. \\
    \texttt{phot\_rp\_mean\_mag} &           12.406217 &                                 Gaia DR2 $G_\mathrm{RP}$ magnitude. \\
       \texttt{radial\_velocity} &           -0.227623 &                              Gaia DR2 heliocentric radial velocity. \\
\texttt{radial\_velocity\_error} &            1.945767 &                               Gaia DR2 radial velocity uncertainty. \\
             \texttt{subcluster} &                halo &                          Is star in core (CG18) or halo (KC19+M21)? \\
               \texttt{in\_CG18} &               False &                             Star in \citet{cantatgaudin_gaia_2018}. \\

@@

             \texttt{(Bp-Rp)\_0} &            0.769225 & Gaia Bp-Rp color, minus $E$($G_\mathrm{BP}$-$G_\mathrm{RP}$)=0.1343 \\
          \texttt{is\_phot\_bin} &                True &                         True if $>0.3$ mag above cluster isochrone. \\
         \texttt{is\_astrm\_bin} &                True &                                                 True if RUWE > 1.2. \\

>>>>

%% This is the title of the table.
\tablecaption{Rotation periods and lithium equivalent widths for
\nkinematic\ candidate NGC\,2516 members.}
\label{tab:maintable}

@@

               \texttt{source\_id} & 5290665204845694336 &                                                                   Gaia DR2 source identifier. \\
         \texttt{source\_id\_edr3} & 5290665204845694336 &                                                                  Gaia EDR3 source identifier. \\
         \texttt{n\_cdips\_sector} &                   7 &                                               Number of TESS sectors with CDIPS light curves. \\
                   \texttt{period} &            0.228981 &                                                              Lomb-Scargle best period [days]. \\
                   \texttt{lspval} &            0.657826 &                                               Lomb-Scargle periodogram value for best period. \\
               \texttt{spdmperiod} &            0.228981 &                                                          Stellingwerf PDM best period [days]. \\
                  \texttt{spdmval} &            0.372779 &                                           Stellingwerf PDM periodogram value for best period. \\
                   \texttt{nequal} &                   0 &                                    Number of stars brighter than the target in TESS aperture. \\
                   \texttt{nclose} &                   1 &                                      Number of stars with $\Delta T > 1.25$ in TESS aperture. \\
                   \texttt{nfaint} &                   1 &                                       Number of stars with $\Delta T > 2.5$ in TESS aperture. \\
                       \texttt{ra} &          119.234778 &                                                               Gaia DR2 right ascension [deg]. \\
                      \texttt{dec} &          -61.015929 &                                                                   Gaia DR2 declination [deg]. \\
               \texttt{ref\_epoch} &              2015.5 &                                          Reference epoch for right ascension and declination. \\
                 \texttt{parallax} &            2.458023 &                                                                      Gaia DR2 parallax [mas]. \\
          \texttt{parallax\_error} &            0.028395 &                                                          Gaia DR2 parallax uncertainty [mas]. \\
                     \texttt{pmra} &           -4.225826 &                           Gaia DR2 proper motion $\mu_\alpha \cos \delta$ [mas$\,$yr${^-1}$]. \\
                    \texttt{pmdec} &           10.899938 &                                       Gaia DR2 proper motion $\mu_\delta$ [mas$\,$yr${^-1}$]. \\
       \texttt{phot\_g\_mean\_mag} &           15.778953 &                                                                       Gaia DR2 $G$ magnitude. \\
      \texttt{phot\_bp\_mean\_mag} &           16.629477 &                                                           Gaia DR2 $G_\mathrm{BP}$ magnitude. \\
      \texttt{phot\_rp\_mean\_mag} &            14.85896 &                                                           Gaia DR2 $G_\mathrm{RP}$ magnitude. \\
         \texttt{radial\_velocity} &                 NaN &                                       Gaia DR2 heliocentric radial velocity [km$\,$s${^-1}$]. \\
  \texttt{radial\_velocity\_error} &                 NaN &                                        Gaia DR2 radial velocity uncertainty [km$\,$s${^-1}$]. \\
               \texttt{subcluster} &                core &                                                    Is star in core (CG18) or halo (KC19+M21)? \\
                 \texttt{in\_CG18} &                True &                                                       Star in \citet{cantatgaudin_gaia_2018}. \\

@@

               \texttt{(Bp-Rp)\_0} &            1.636198 & Gaia $G_\mathrm{BP}$-$G_\mathrm{RP}$ color, minus $E$($G_\mathrm{BP}$-$G_\mathrm{RP}$)=0.1343 \\
            \texttt{is\_phot\_bin} &               False &                                                   True if $>0.3$ mag above cluster isochrone. \\
           \texttt{is\_astrm\_bin} &               False &                                                                 True if Gaia EDR3 RUWE > 1.2. \\
                     \texttt{ruwe} &            0.986349 &                                                                               Gaia EDR3 RUWE. \\
      \texttt{Li\_EW\_mA\_GaiaESO} &              -1.495 &                          Gaia-ESO Li doublet equivalent width, including the Fe blend [m\AA]. \\
\texttt{Li\_EW\_mA\_perr\_GaiaESO} &              12.407 &                                     Gaia-ESO Li doublet EW upper uncertainty interval [m\AA]. \\
\texttt{Li\_EW\_mA\_merr\_GaiaESO} &                 5.0 &                                     Gaia-ESO Li doublet EW lower uncertainty interval [m\AA]. \\
        \texttt{Li\_EW\_mA\_GALAH} &                 NaN &                             GALAH Li doublet equivalent width, including the Fe blend [m\AA]. \\
  \texttt{Li\_EW\_mA\_perr\_GALAH} &                 NaN &                                        GALAH Li doublet EW upper uncertainty interval [m\AA]. \\
  \texttt{Li\_EW\_mA\_merr\_GALAH} &                 NaN &                                        GALAH Li doublet EW lower uncertainty interval [m\AA]. \\

### 2021-04-26 cluster_paper/ms.tex
update ms

\keywords{
	stellar ages (1581),
  stellar associations (1582),
  open clusters (1160),
  stellar kinematics (1608),
  stellar rotation (1629)
}

@@

The third and final approach we took for confirming the youth of the
halo population of \cn\ was an analysis of the Li\,\textsc{I}
6708\,\AA\ doublet.  This type of age measurement has been reviewed by
\citet{soderblom_ages_2010}, and its connection to stellar rotation
has recently been surveyed by \citet{bouvier_lithium-rotation_2020}.
For \cn, two spectroscopic datasets seemed important:
Gaia-ESO \citep{gilmore_gaiaeso_2012} and GALAH
\citep{silva_galah_2015}.  At the time of analysis, Gaia-ESO DR4 and
GALAH DR3 were the most relevant, respectively \citep[{\it
e.g.},][]{randich_gaiaeso_2018,buder_galah_2020}.  The target
selection and results from each survey were as follows.

The Gaia-ESO collaboration chose candidate \cn\ members to observe
with GIRAFFE and UVES based on previously reported literature members
and publicly available photometry.  Since the existence of the \cn\
halo was not known at the time of target selection, very few halo
stars are in the sample.  For \cn, \citet{randich_gaiaeso_2018}
reported stellar parameters (including lithium equivalent widths and
metallicities) for 796 stars that they considered possible \cn\
members.  Cross-matching these against our kinematic list of
\nkinematic\ candidate members by position and imposing a 0$\farcs$5
maximum separation limit yielded 459 kinematic members with available
spectra, 417 in the core and 42 in the halo of \cn.  The lop-sided
ratio is due to the Gaia-ESO selection function.  An inspection of the
effective temperature {\it vs{.}} lithium equivalent width based on
their Table~2 showed the surprising feature that stars with $T_{\rm
eff}\lesssim 4000\,{\rm K}$ tend to have non-zero equivalent widths,
despite the expectation that their lithium should be fully depleted.
Whether these measurements were significant was not clear, and so we
opted to download the spectra from the ESO archive to re-measure
equivalent widths.

The GALAH DR3 target selection is discussed in depth by
\citet{buder_galah_2020}.  The relevant aspects for our analysis are
that the ``main'' survey targeted $12<V<14$ stars at
$\delta<+10^\circ$, provided the stars were at least ten degrees from
the galactic plane.  The survey is spatially
inhomogeneous.\footnote{See the footprint at
\url{https://www.galah-survey.org/news/announcing-galah-dr3}} Special
targeting for stars in the TESS southern continuous viewing zone, and
for known open cluster members was also performed.  We identified the
candidate \cn\ members for which spectra had been obtained by
searching the \texttt{GALAH\_DR3\_main\_allstar\_v1} catalog, after
excluding stars with the stellar parameter bit flags 1, 2, 3.  This
excludes spectra with unreliable broadening, low S/N, and unreliable
wavelength solutions (see Table~6 of \citealt{buder_galah_2020}).
Since our main focus is measuring equivalent widths for the
Li\,6708\,\AA\ doublet, these were the most relevant bit flags.  Of
our \nkinematic\ candidate \cn\ members, 107 had spectra in GALAH DR3.
51 were in the ``core''; 56 were in the ``halo''.  78 had ``finite''
lithium, i.e., no lithium flag set.
We downloaded\footnote{Via \url{datacentral.org.au/services/download},
using the \texttt{sobject\_id} identifiers.} the GALAH DR3 spectra for
all 107 entries.

@@

We measured the lithium equivalent widths using the \texttt{specutils}
package \citep{specutils_v1pt1}.  We did this by first focusing in on
a 15\AA\ window centered on the 6707.835\AA\ lithium doublet.  We then
continuum normalized the spectra using a third-degree Chebyshev
series, while excluding any regions that showed absorption lines ({\it
e.g.}, Fe\,\textsc{I} is present at 6703.58\,\AA and 6705.10\,\AA).
We proceeding by fitting a Gaussian to the continuum normalized
spectra, considering only a $\pm 1\,$\AA\ window centered on the Li
doublet.  The equivalent widths were then evaluated by numerically
integrating the fitted model over the same window.  Our approach
therefore includes the Fe\,6707.44 blend in the reported Li equivalent
widths---which leads to systematic overestimates of $\approx$10 to
15\,m\AA\ \citep[{\it e.g.},][]{bouvier_pleiades_lirot_2018}.
Finally, to derive uncertainties on the EWs, we repeated the procedure
twenty times, but added noise to the spectra, drawn from a normal
distribution with a scale set by the standard deviation of the
continuum.  The reported uncertainties are then drawn from the
$84^{\rm th}$ and $16^{\rm th}$ percentiles of the resulting EW
distribution.  We verified the overall scale of our results by
comparing our measurements with those of \citet{randich_gaiaeso_2018}.
Most stars with $T_{\rm eff} > 4500\,{\rm K}$ followed a 1-1 relation.
For cooler stars, our EW measurements are systematically lower, as
expected given the depletion timescales for stars at these ages
\citep[{\it e.g.},][]{soderblom_ages_2014}.  We visually verified from
the spectra that, for the most part, we prefer our measurements.

@@

Any differences in the stellar mass function between the core and halo
could be informative, since the tidal tail explanation predicts that
the mean tail star is expected to be less massive than the mean core
star \citep[{\it e.g.},][]{chumak_tails_2006}.
Figure~\ref{fig:massfunction} shows the histogram of $A_G$-corrected
absolute magnitudes, with the corresponding
\citet{pecaut_mamajek_2013} mass estimates.  The stars closest
to the core do
indeed seem to have a greater proportion of the $\gtrsim$0.8$M_\odot$
stars, while the tail stars have a greater proportion of
$\lesssim$0.4$M_\odot$ stars.  The mean star in the inner separation bin has
$({\rm M}_G)_0=7.08$ ($\approx$0.69$M_\odot$), while in the outer bin
the mean star has $({\rm M}_G)_0=7.61$ ($\approx$0.63$M_\odot$).
While the difference does appear significant given the number of
stars involved, the theoretical implications are not entirely clear,
since the ``molecular filament'' explanation might be able to
accommodate mass differences between the core and tail through density
gradients across the initial filament.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f15.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Absolute magnitudes (and masses) for the inner and
			outermost stars in \cn.}
		The \nkinematic\ candidate \cn\ members were split into three groups
		according to their distance from the cluster center.
		The inner and outermost groups, each containing 1,099 stars, are shown to
		provide the clearest picture with respect to expectations for mass
		segregation.
		The mean halo star does have lower mass, consistent with expectations
		for tidal tails.
		\label{fig:massfunction}
	}
\end{figure*}

@@

   \item {\it The average star in the halo has lower mass (by $\approx$10\%)
   than the average star in the core.}

@@

The existence of the halo itself is physically plausible.  A
star with a 1\kms\ velocity difference from the cluster average will
move away from the center at a rate of 1\,pc\,Myr$^{-1}$.  For a
$\sim$100\,Myr cluster, a halo with characteristic size $\sim$100\,pc
is therefore expected, given that the typical velocity dispersions of
open clusters are of order kilometers per second.

On the other hand, our ability to reliably identify members of these
halos is rather surprising.  These stars have remained hidden in the
background of the Galaxy throughout centuries of modern astronomy.
The consequences that these halo stars will have in expanding our
understanding of planetary and stellar origins are only just beginning
to be understood.

>>>>

\keywords{
  stellar associations (1582),
  open clusters (1160),
	stellar ages (1581),
  stellar kinematics (1608),
  stellar rotation (1629)
}

@@

The third and final approach we took for confirming the youth of the
halo population of \cn\ was an analysis of the Li\,\textsc{I}
6708\,\AA\ doublet.  This type of age measurement has been reviewed by
\citet{soderblom_ages_2010}, and its connection to stellar rotation
has recently been surveyed by \citet{bouvier_lithium-rotation_2020}.
For \cn, two spectroscopic datasets seemed important:
Gaia-ESO \citep{gilmore_gaiaeso_2012} and GALAH
\citep{silva_galah_2015}.  At the time of analysis, Gaia-ESO DR4 and
GALAH DR3 were the most relevant
\citep{randich_gaiaeso_2018,buder_galah_2020}.  The target selection
and results from each survey were as follows.

The Gaia-ESO collaboration chose candidate \cn\ members to observe
with GIRAFFE and UVES based on previously reported literature members
and publicly available photometry.  Since the existence of the \cn\
halo was not known at the time of target selection, very few halo
stars are in the sample.  Based on the data,
\citet{randich_gaiaeso_2018} determined stellar parameters (including
lithium equivalent widths and metallicities) for 796 stars that they
considered possible \cn\ members.  Cross-matching these against our
kinematic list of \nkinematic\ candidate members by position and
imposing a 0$\farcs$5 maximum separation limit yielded 459 kinematic
members with available spectra, 417 in the core and 42 in the halo of
\cn.  The lop-sided ratio is due to the Gaia-ESO selection function.
An inspection of the effective temperature {\it vs{.}} lithium
equivalent width based on Table~2 of \citet{randich_gaiaeso_2018}
showed the surprising feature that stars with $T_{\rm eff}\lesssim
4000\,{\rm K}$ tend to have non-zero equivalent widths, despite the
expectation that their lithium should be fully depleted.  Whether
these measurements were significant was not clear, and so we opted to
download the spectra from the ESO archive to re-measure equivalent
widths.

The GALAH DR3 target selection is discussed in depth by
\citet{buder_galah_2020}.  The relevant aspects for our analysis are
that the ``main'' survey targeted $12<V<14$ stars at
$\delta<+10^\circ$ in select fields, provided the stars were at least
ten degrees from the galactic plane.  Special targeting for stars in
the TESS southern continuous viewing zone, and for known open cluster
members was also performed.  We identified the candidate \cn\ members
for which spectra had been obtained by searching the
\texttt{GALAH\_DR3\_main\_allstar\_v1} catalog, after excluding stars
with the stellar parameter bit flags 1, 2, 3.  This excludes spectra
with unreliable broadening, low S/N, and unreliable wavelength
solutions (see Table~6 of \citealt{buder_galah_2020}).
% Since our main focus is measuring equivalent widths for the
% Li\,6708\,\AA\ doublet, these were the most relevant bit flags.
Of our \nkinematic\ candidate \cn\ members, 107 had spectra in GALAH
DR3.  51 were in the core, and 56 were in the ``halo''.  We
downloaded\footnote{Via \url{datacentral.org.au/services/download},
using the \texttt{sobject\_id} identifiers.} the GALAH DR3 spectra for
all 107 entries.

@@

We measured the lithium equivalent widths using the \texttt{specutils}
package \citep{specutils_v1pt1}.  We did this by first shifting the
spectra to the stellar rest frames, and then focused in on a 15\AA\
window centered on the 6707.835\AA\ lithium doublet.  We then
continuum normalized the spectra using a third-degree Chebyshev
series, while excluding any regions that showed absorption lines ({\it
e.g.}, Fe\,\textsc{I} is present at 6703.58\,\AA\ and 6705.10\,\AA).
We proceeding by fitting a Gaussian to the continuum normalized
spectra, considering only a $\pm 1\,$\AA\ window centered on the Li
doublet.  The equivalent widths were then evaluated by numerically
integrating the fitted model over the same window.  Our approach
therefore includes the Fe\,6707.44\AA\ blend in the reported Li equivalent
widths---which leads to systematic overestimates of $\approx$10 to
15\,m\AA\ \citep[{\it e.g.},][]{bouvier_pleiades_lirot_2018}.
Finally, to derive uncertainties on the EWs, we repeated the procedure
twenty times, but added noise to the spectra, drawn from a normal
distribution with a scale set by the standard deviation of the
continuum.  The reported uncertainties are then drawn from the
$84^{\rm th}$ and $16^{\rm th}$ percentiles of the resulting EW
distribution, and are included in Table~\ref{tab:maintable}.  We
verified the overall scale of our results by comparing our
measurements with those of \citet{randich_gaiaeso_2018}.  Most stars
with $T_{\rm eff} > 4500\,{\rm K}$ followed a 1-1 relation.  For
cooler stars, our EW measurements are systematically lower, as
expected given the depletion timescales for stars at these ages
\citep[{\it e.g.},][]{soderblom_ages_2014}.  We visually verified from
the spectra that, for the most part, we prefer our measurements.

@@

Any differences in the stellar mass function between the core and halo
could be informative, since the tidal tail explanation predicts that
the mean tail star is expected to be less massive than the mean core
star \citep[{\it e.g.},][]{chumak_tails_2006}.
Figure~\ref{fig:massfunction} shows the histogram of $A_G$-corrected
absolute magnitudes, with the corresponding
\citet{pecaut_mamajek_2013} mass estimates.  The stars closest to the
cluster center do seem to have a greater proportion of $\gtrsim$0.8$M_\odot$
stars, while the outermost stars have a greater proportion of
$\lesssim$0.4$M_\odot$ stars.  The mean star in the inner separation
bin has $({\rm M}_G)_0=7.08$ ($\approx$0.69$M_\odot$), while in the
outer bin the mean star has $({\rm M}_G)_0=7.61$
($\approx$0.63$M_\odot$).  While the difference does appear
significant given the number of stars involved, the theoretical
implications are not entirely clear, since the ``molecular filament''
explanation might be able to accommodate mass differences between the
core and tail through density gradients across the initial filament.
Observational incompleteness at the low-mass end seems unlikely to
alter the result though, since the innermost and outermost are likely
subject to similar selection effects.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f15.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Absolute magnitudes and masses for the inner and
  outermost stars in \cn.} The \nkinematic\ candidate \cn\ members
  were split into three groups according to their distance from the
  cluster center.  The inner and outermost groups, each containing
  1,099 stars, are shown in 1\,mag bins.  The mean star in the
  innermost sample has a 10\% lower mass than the mean star in the
  outermost sample, consistent with expectations for tidal tails.
  \label{fig:massfunction}
	}
\end{figure*}

@@

   \item {\it The average star in the outskirts has lower mass (by $\approx$10\%)
   than the average star in the cluster center.}

@@

The existence of the halo itself is not particularly surprising.  A
star with a 1\kms\ velocity difference from the cluster average will
move away from the center at a rate of 1\,pc\,Myr$^{-1}$.  For a
$\sim$100\,Myr cluster, a halo with characteristic size $\sim$100\,pc
is therefore expected, given that the typical velocity dispersions of
open clusters are of order kilometers per second.

On the other hand, our ability to reliably identify members of these
halos is rather surprising.  These stars have remained hidden in the
background of the Galaxy throughout centuries of modern astronomy.
Their discovery likely represents a major step toward understanding
open cluster dispersal.
Combined with wide-field photometric surveys, the stars themselves
will enable advances in understanding stellar
angular momentum evolution.
Combined with wide-field spectroscopic
surveys \citep[{\it e.g.},][]{kollmeier_2017}, the halos could also lead to
advances in understanding the chemical evolution of low-mass stars.
Finally, the halo stars also provide an expanded sample of
age-dated stars to search for planets \citep[{\it
e.g.},][]{newton_2021}, which could help expand our understanding of
planetary evolution.
